The importance of transient resonances in extreme-mass-ratio inspirals

### 2016-08-31 res_char.tex
First comments from Stas

Resonances involving the azimuthal motion do not produce a comparable effect because of the axisymmetry of the background spacetime. However, both $\theta$--$\phi$ resonances~\cite{Hirata2011} and $r$--$\phi$ resonances~\cite{VanDeMeent2013} can lead to extrinsic effects; the GWs from such systems are not emitted isotropically and the imbalance produces a kick velocity that is, in some cases, sufficient to eject the central BH from its host~\cite{VanDeMeent2014}.

@@

To follow the evolution of the inspiral we must have a means of prescribing the forcing acceleration which causes the orbit to deviate from a single geodesic. We work directly with the gravitational self-force, using the same post-Newtonian (PN) approximation as Flanagan and Hinderer~\cite{Flanagan2012}. For comparison, Flanagan, Hughes and Ruangeri~\cite{Flanagan2012a} use a Teukolsky-equation calculation of GW fluxes to account for the inspiral due to radiation reaction.

@@

We also use the resonance timescale (cf.\ \cite{Ruangsri2014})
\begin{equation}
\tau\sub{res} = \left[\frac{2\pi}{\left|\left\langle\dot{\Omega}(0)\right\rangle_{q'}\right|}\right]^{1/2}.
\label{eq:t-res}
\end{equation}
Here $\dot{\Omega}(0)$ is the rate of change of $\Omega$ at resonance, which we take to be at $t = 0$. The instantaneous $\dot{\Omega}$ depends upon the orbital phase and oscillates about its mean trend over an orbit. We are interested in the averaged behaviour, not the periodic modulations about this, which is why we use the time-average $\langle\dot{\Omega}\rangle_{q'}$; here we use $q'$ to represent a phase that varies over an orbit with period of order $T$. Close to resonance, $\Omega(t)$ is well approximated by a first-order Taylor expansion, decreasing linearly with time; hence we make the approximation
\begin{equation}
\left|{\Omega(t)}\right| \simeq \left|\left\langle\dot{\Omega}(0)\right\rangle_{q'} t\right|.
\label{eq:Taylor-Omega}
\end{equation}
The resonant timescale should give an indication of the time over which we expect the effects of the resonance to be felt~\cite{Bosley1992}. Consider the phase of the Mino-time Fourier expansion on resonance; neglecting the constant, the resonant Fourier component has form
\begin{equation}
\varphi_{n_r,\,-n_\theta} \simeq \left(n_r\Upsilon_r - n_\theta\Upsilon_\theta\right)\lambda + \left(n_r\dot{\Upsilon}_r - n_\theta\dot{\Upsilon}_\theta\right)\lambda^2 + \ldots
\end{equation}
Typically, the first term is non-zero and this gives the familiar oscillation. On resonance, it is zero, leaving the next-order term to govern the behaviour~\cite{Flanagan2012,Ruangsri2014}. Only once we have moved far enough away from resonance for the first term to dominate the second do we recapture the non-resonant behaviour. The first term (translating from Mino time to coordinate time) sets $\tau\sub{pres}$, the second sets $\tau\sub{res}$.

@@

Evolving through a resonance can lead to an enhancement (or decrement) of fluxes relative to the adiabatic prescription. After crossing the resonance region, the orbital parameters are different from those calculated from an adiabatic evolution. Flanagan and Hinderer~\cite{Flanagan2012} gave an expression for this deviation. If we denote the orbital parameters by $\mathcal{I}^a = \{E,L_z,Q\}$, then the change across resonance is
\begin{align}
\Delta \mathcal{I}^a = {} & \eta\sum_{s\,\neq\,0}F_{a,\,s}^{(1)}\left[\dfrac{2\pi}{\left|s \left\langle\dot{\Omega}\right\rangle_{q'}\right|}\right]^{1/2} \nonumber \\*
 {} & \times {} \exp\left[i\left(s \widehat{\kappa}_0 + \dfrac{\pi}{4} \sgn s\dot{\Omega}\right)\right]. 
\end{align}
Here $\widehat{\kappa}_0$ is the orbital phase on resonance and $F_{a,\,s}^{(1)}$ is the $s$-th harmonic of the first-order self-force on resonance, defined such that\footnote{Since the geodesic equations decouple in Mino time rather than coordinate time, this is true only in an average sense.}
\begin{equation}
\diff{\mathcal{I}^a}{t} = \eta\sum_s F_{a,\,s}^{(1)}(\boldsymbol{\mathcal{I}})\exp(is q) + \order{\eta^2}.
\end{equation}
A derivation is presented in \apref{res-asymptotic}, which contains a more comprehensive explanation of the various terms. This employs matched asymptotic expansions to track the evolution through resonance, following the approach of Kevorkian~\cite{Kevorkian1987}.

@@

We wish to test whether there exists some set of parameters $\boldsymbol{\Theta}$ such that the resulting adiabatic waveform is sufficiently similar to the full waveform. We do this by evaluating the SNR for each (normalised) waveform template,
\begin{equation}
\rho\left[h\right] = \max_t \frac{\overlap{s}{h}}{\sqrt{\overlap{h}{h}}}.
\end{equation}
We maximise over the time offset for the template to find the best fit to the data. If a template exactly matches the data, it would produce an SNR of $\sqrt{\overlap{s}{s}}$, hence the overlap
\begin{equation}
\label{eq:overlap}
\mathbb{O}\left[h\right] = \frac{\overlap{s}{h}}{\sqrt{\overlap{s}{s}}\sqrt{\overlap{h}{h}}},
\end{equation}
which ranges from $0$ to $1$ provides a convenient indication of how well-matched the template is to the data.

@@

Before investigating the impact of resonances on an EMRI signal, we first compare results from the full instantaneous evolution and the $2$-torus averaged adiabatic evolution over $2~\mathrm{yr}$ for an inspiral which avoids any significant resonances. Example evolutions of the orbital parameters $E$, $L_z$ and $Q$ are shown in \figref{good-traj}, which shows that the adiabatic evolution closely matches the full evolution on longer inspiral timescales. The inset plots show the start of the evolution, on a timescale associated with the orbital motion of the compact object; the $2$-torus averaging explicitly smooths out the visible structure on this scale. The two appraches are in good agreement.

@@

\begin{figure}
\centering
%\includegraphics[width=0.46\textwidth]{Fig_overlap_vs_time}
\includegraphics[width=0.4\textwidth]{Fig_overlap}
\caption{\label{fig:overlap-dephasing}The overlap computed between the full evolution and an adiabatic evolution for an illustrative EMRI system with $p=7.85$, as a function of time, including only the parts of the waveforms within some $\Delta t$ of $t$, arbitrarily chosen to give $25$ independent (non-overlapping) calculations. The time $t\sub{res} = 0$ is when the full evolution crosses the $2$:$3$ resonance.}
\end{figure}

@@

We have computed this family of adiabatic evolutions for our illustrative resonance. In \figref{res-diff-traj}, we plot the difference in the orbital parameters ($E$, $L_z$ and $Q$) between the various trajectories and the adiabatic evolution that matches at the start. The jumps in the orbital parameters due to the $2$:$3$ resonance can be clearly seen, as can the fast orbital oscillations present in the full instanteous evolution but absent in the adiabatic evolutions.

@@

Let us now consider jumps for systems other than our illustrative $2$:$3$ resonance. We expect nearly-circular orbits to encounter smaller jumps than more eccentric orbits~\cite{Flanagan2012} because they have a smaller $r$--$\theta$ $2$-torus, meaning that resonant orbits come closer to every allowed point, nearer approximating a non-resonant orbit. In \figref{res-flux-rms-p}, we show the root-mean-square (over a grid of $a_\ast$ and $\cos\iota$ values) relative resonant jump for the semilatus rectum $p$ as a function of eccentricity $e$ for a selection of low-order resonances. The other orbital shape parameters show a similar trend with increasing eccentircity~\cite{ColeThesis2015}. Larger eccentricities do give rise to larger jumps, matching our expectations and Teukolsky-based calculations by Flanagan, Hughes and Ruangsri~\cite{Flanagan2012a}.

@@

Roughly $30\%$ of the systems lie within $0.05$ of the expected value. However, the vast majority of these are not significantly disrupted by resonances, and produce overlaps approaching unity. For the smallest (most extreme) mass ratios, the inspiral rate is slow, and so the systems do not encounter either the $1$:$2$ or $2$:$3$ resonances during their lifetime. Meanwhile, for the largest (least extreme) mass ratios, the EMRIs encounter both resonances close to plunge. In each case, there is a long resonance-free region, allowing a high overlap to be recovered.

The remaining $66\%$ of systems have overlaps above the level expected if resonances lead to significant dephasing. These occur at low and intermediate values of the mass ratio, where the inspiral rate is slow enough that the low-order resonances are encountered in the middle of the observation window, and the resulting value of $t\sub{ad}$ is small. For these EMRIs, resonances are not as important as expected, which is likely due to the low eccentricities of our population leading to small (and in most cases, negligible) resonant flux enhancements.

@@

\begin{acknowledgments}
The authors extend our sincere gratitude to \'{E}anna Flanagan and Tanja Hinderer for providing their PN self-force code, without which this work would not be possible. We thank Tanja Hinderer, Jeandrew Brink, Maarten van de Meent, Leor Barack, Scott Hughes and Nico Yunes for useful conversations; Stanislav Babak for examining both \cite{BerryThesis2013} and \cite{ColeThesis2015}, and Christopher Moore with help preparing this manuscript. RHC was supported by STFC; CPLB thanks STFC and the Cambridge Philosophical Society; PC's work was supported by a Marie Curie Intra-European Fellowship within the 7th European Community Framework Programme (PIEF-GA-2011-299190), and JRG was supported by the Royal Society. This is LIGO Document P1600251. 
\end{acknowledgments}

>>>>

Resonances involving the azimuthal motion do not produce a comparable effect because of the axisymmetry of the background spacetime. However, both $\theta$--$\phi$ resonances~\cite{Hirata2011} and $r$--$\phi$ resonances~\cite{VanDeMeent2013} can lead to extrinsic effects; the GWs from such systems are not emitted isotropically and the imbalance produces a kick velocity that is, in some cases with moderate mass ratios, sufficient to eject the central BH from its host~\cite{VanDeMeent2014}.

@@

To follow the evolution of the inspiral we must have a means of prescribing the forcing acceleration which causes the orbit to deviate from a single geodesic. We work directly with the gravitational self-force, using the same post-Newtonian (PN) approximation as Flanagan and Hinderer~\cite{Flanagan2012}. For comparison, Flanagan, Hughes and Ruangsri~\cite{Flanagan2012a} use a Teukolsky-equation calculation of GW fluxes to account for the inspiral due to radiation reaction.

@@

We also use the resonance timescale (cf.\ \cite{Ruangsri2014})
\begin{equation}
\tau\sub{res} = \left[\frac{2\pi}{\left|\left\langle\dot{\Omega}(0)\right\rangle_{q'}\right|}\right]^{1/2}.
\label{eq:t-res}
\end{equation}
Here $\dot{\Omega}(0)$ is the rate of change of $\Omega$ at resonance, which we take to be at $t = 0$. The instantaneous $\dot{\Omega}$ depends upon the orbital phase and oscillates about its mean trend over an orbit. We are interested in the averaged behaviour, not the periodic modulations about this, which is why we use the time-average $\langle\dot{\Omega}\rangle_{q'}$; here we use $q'$ to represent a phase that varies over an orbit with period of order $T$.\footnote{On resonance, we are interested in the relative $r$--$\theta$ orbital phase ($\psi_{\theta_-}$ or equivalent), which sets the resonant trajectory in the $r$--$\theta$ plane, but not the exact phase of the orbit around this loop.} Close to resonance, $\Omega(t)$ is well approximated by a first-order Taylor expansion, decreasing linearly with time; hence we make the approximation
\begin{equation}
\left|{\Omega(t)}\right| \simeq \left|\left\langle\dot{\Omega}(0)\right\rangle_{q'} t\right|.
\label{eq:Taylor-Omega}
\end{equation}
The resonant timescale should give an indication of the time over which we expect the effects of the resonance to be felt~\cite{Bosley1992}. Consider the phase of the Mino-time Fourier expansion on resonance; neglecting the constant, the resonant Fourier component has form
\begin{equation}
\varphi_{n_r,\,-n_\theta} \simeq \left(n_r\Upsilon_r - n_\theta\Upsilon_\theta\right)\lambda + \left(n_r\dot{\Upsilon}_r - n_\theta\dot{\Upsilon}_\theta\right)\lambda^2 + \ldots
\end{equation}
Typically, the first term is non-zero and this gives the familiar oscillation. On resonance, it is zero, leaving the next-order term to govern the behaviour~\cite{Flanagan2012,Ruangsri2014}. Only once we have moved far enough away from resonance for the first term to dominate the second do we recapture the non-resonant behaviour. The first term (translating from Mino time to coordinate time) sets $\tau\sub{pres}$, the second sets $\tau\sub{res}$.

@@

Evolving through a resonance can lead to an enhancement (or decrement) of fluxes relative to the adiabatic prescription. After crossing the resonance region, the orbital parameters are different from those calculated from an adiabatic evolution. Flanagan and Hinderer~\cite{Flanagan2012} gave an expression for this deviation. If we denote the orbital parameters by $\mathcal{I}^a = \{E,L_z,Q\}$, then the change across resonance is
\begin{align}
\Delta \mathcal{I}^a = {} & \eta\sum_{s\,\neq\,0}F_{a,\,s}^{(1)}\left[\dfrac{2\pi}{\left|s \left\langle\dot{\Omega}\right\rangle_{q'}\right|}\right]^{1/2} \nonumber \\*
 {} & \times {} \exp\left[i\left(s \widehat{\kappa}_0 + \dfrac{\pi}{4} \sgn s\dot{\Omega}\right)\right]. 
\end{align}
Here $\widehat{\kappa}_0$ is the phase on resonance, which sets the resonant trajectory in the $r$--$\theta$ plane similarly to $\psi_{\theta_-}$, and $F_{a,\,s}^{(1)}$ is the $s$-th harmonic of the first-order self-force on resonance, defined such that\footnote{Since the geodesic equations decouple in Mino time rather than coordinate time, this is true only in an average sense.}
\begin{equation}
\diff{\mathcal{I}^a}{t} = \eta\sum_s F_{a,\,s}^{(1)}(\boldsymbol{\mathcal{I}})\exp(is q) + \order{\eta^2}.
\end{equation}
A derivation is presented in \apref{res-asymptotic}, which contains a more comprehensive explanation of the various terms. This employs matched asymptotic expansions to track the evolution through resonance, following the approach of Kevorkian~\cite{Kevorkian1987}.

@@

We wish to test whether there exists some set of parameters $\boldsymbol{\Theta}$ such that the resulting adiabatic waveform is sufficiently similar to the full waveform. We do this by evaluating the SNR for each (normalised) waveform template,
\begin{equation}
\rho\left[h\right] = \max_t \frac{\overlap{s}{h}}{\sqrt{\overlap{h}{h}}}.
\end{equation}
We maximise over the time offset for the template to find the best fit to the data. If a template exactly matches the data, it would produce an SNR of $\sqrt{\overlap{s}{s}}$, hence the overlap
\begin{equation}
\label{eq:overlap}
\mathbb{O}\left[h\right] = \max_t \frac{\overlap{s}{h}}{\sqrt{\overlap{s}{s}}\sqrt{\overlap{h}{h}}},
\end{equation}
which ranges from $0$ to $1$ provides a convenient indication of how well-matched the template is to the data.

@@

Before investigating the impact of resonances on an EMRI signal, we first compare results from the full instantaneous evolution and the $2$-torus averaged adiabatic evolution over $2~\mathrm{yr}$ for an inspiral which avoids any significant resonances. Example evolutions of the orbital parameters $E$, $L_z$ and $Q$ are shown in \figref{good-traj}, which shows that the adiabatic evolution closely matches the full evolution on longer inspiral timescales. The inset plots show the start of the evolution, on a timescale associated with the orbital motion of the compact object; the $2$-torus averaging explicitly smooths out the visible structure on this scale. The two approaches are in good agreement.

@@

\begin{figure}
\centering
%\includegraphics[width=0.46\textwidth]{Fig_overlap_vs_time}
\includegraphics[width=0.4\textwidth]{Fig_overlap}
\caption{\label{fig:overlap-dephasing}The overlap computed between the full evolution and an adiabatic evolution for an illustrative EMRI system with $p=7.85$, as a function of time, including only the parts of the waveforms within some $\Delta t$ of $t$, arbitrarily chosen to give $25$ independent (non-overlapping) calculations. The time $t = t\sub{res}$ is when the full evolution crosses the $2$:$3$ resonance.}
\end{figure}

@@

We have computed this family of adiabatic evolutions for our illustrative resonance. In \figref{res-diff-traj}, we plot the difference in the orbital parameters ($E$, $L_z$ and $Q$) between the various trajectories and the adiabatic evolution that matches at the start. The jumps in the orbital parameters due to the $2$:$3$ resonance can be clearly seen, as can the fast orbital oscillations present in the full instantaneous evolution but absent in the adiabatic evolutions.

@@

Let us now consider jumps for systems other than our illustrative $2$:$3$ resonance. We expect nearly-circular orbits to encounter smaller jumps than more eccentric orbits~\cite{Flanagan2012} because they have a smaller $r$--$\theta$ $2$-torus, meaning that resonant orbits come closer to every allowed point, nearer approximating a non-resonant orbit. In \figref{res-flux-rms-p}, we show the root-mean-square (over a grid of $a_\ast$ and $\cos\iota$ values) relative resonant jump for the semilatus rectum $p$ as a function of eccentricity $e$ for a selection of low-order resonances. The other orbital shape parameters show a similar trend with increasing eccentricity~\cite{ColeThesis2015}. Larger eccentricities do give rise to larger jumps, matching our expectations and Teukolsky-based calculations by Flanagan, Hughes and Ruangsri~\cite{Flanagan2012a}.

@@

Roughly $30\%$ of the systems lie within $0.05$ of the expected value. The vast majority of these are not significantly disrupted by resonances, and produce overlaps approaching unity. For the smallest (most extreme) mass ratios, the inspiral rate is slow, and so the systems do not encounter either the $1$:$2$ or $2$:$3$ resonances during their lifetime. Meanwhile, for the largest (least extreme) mass ratios, the EMRIs encounter both resonances close to plunge. In each case, there is a long resonance-free region, allowing a high overlap to be recovered.

The remaining $66\%$ of systems have overlaps above the level expected if resonances lead to significant dephasing. These occur at low and intermediate values of the mass ratio, where the inspiral rate is slow enough that the low-order resonances are encountered in the middle of the observation window, and the resulting value of $t\sub{ad}$ is small. For these EMRIs, resonances are not as important as expected. Some of these could be because of fortuitous phases on resonance corresponding to small resonance jumps. The most likely explanation for the lack of impact is because of the low eccentricities of our population, this means that the magnitude of the resonant flux enhancements are small (and in most cases negligible).

@@

\begin{acknowledgments}
The authors extend our sincere gratitude to \'{E}anna Flanagan and Tanja Hinderer for providing their PN self-force code, without which this work would not be possible. We are grateful to Stanislav Babak for examining both \cite{BerryThesis2013} and \cite{ColeThesis2015}, and for providing comments on this draft. We thank Tanja Hinderer, Jeandrew Brink, Maarten van de Meent, Leor Barack, Scott Hughes and Nico Yunes for useful conversations, and Christopher Moore with help preparing this manuscript. RHC was supported by STFC; CPLB thanks STFC and the Cambridge Philosophical Society; PC's work was supported by a Marie Curie Intra-European Fellowship within the 7th European Community Framework Programme (PIEF-GA-2011-299190), and JRG was supported by the Royal Society. This is LIGO Document P1600251. 
\end{acknowledgments}
